---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data

### Load the data

Data loaded into **activity** tibble using **tidyverse::read_csv()** 

```{r load, echo=TRUE, message=FALSE}
library(tidyverse)
activity <- read_csv("activity.zip")
```

## What is mean total number of steps taken per day?

### Histogram of total number of steps taken per day

```{r stepshist, echo=TRUE}
activity %>% drop_na(steps) %>%
      group_by(date) %>% 
      summarise(daily_steps = sum(steps, na.rm = T)) %>% 
      ggplot(aes(x = daily_steps)) +
      geom_histogram(binwidth = 2000,
                     fill = "steelblue",
                     colour = "white") +
      labs(title = "Daily steps histogram",
           x = "Daily steps",
           y = "") +
      theme_minimal()
```

### Mean and median of total number of steps taken per day

```{r stepsmed, echo=TRUE}
daily_steps <- activity %>% drop_na(steps) %>%
      group_by(date) %>% 
      summarise(steps = sum(steps, na.rm = T))
options(scipen = 999999)
daily_steps_mean <- mean(daily_steps$steps)
daily_steps_median <- median(daily_steps$steps)
```

The mean of total number of steps taken per day is **`r round(daily_steps_mean, 2)`** and the median is **`r round(daily_steps_median, 2)`**.

## What is the average daily activity pattern?

### Time series plot of average steps across intervals

```{r dailypattern, echo=TRUE}
activity %>% drop_na(steps) %>%
      group_by(interval) %>%
      summarise(interval_mean_steps = mean(steps, na.rm = T)) %>% 
      ggplot(aes(x = interval, y = interval_mean_steps)) +
      geom_line(colour = "steelblue") +
      scale_x_continuous(breaks = seq(0, 2300, by = 200)) +
      labs(title = "Average steps for each interval",
           x = "Interval",
           y = "Average steps") +
      theme_minimal()
```

### Interval with highest average number of steps

```{r maxavgsteps, echo=TRUE}
interval_max_steps <- activity %>% drop_na(steps) %>%
      group_by(interval) %>%
      summarise(interval_mean_steps = mean(steps, na.rm = T)) %>% 
      filter(interval_mean_steps == max(interval_mean_steps))
```

5-minute interval that, on average across all the days in the dataset, contains the maximum number of steps is number **`r interval_max_steps$interval`**.


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
